\documentclass[../RFC-HDFG-2026-003.tex]{subfiles}

\begin{document}

\section{Filter Plugin Class Extension}
\label{sec:class-extension}

\subsection{Blob Callbacks}
\label{sec:blob-callbacks}

Three new, NULL-able callbacks are added to \texttt{H5Z\_class3\_t}
(RFC-HDFG-2026-001): \texttt{write\_blob}, \texttt{read\_blob}, and
\texttt{close\_blob}. When NULL, the library uses a default
implementation built on \texttt{H5HG} (the existing global-heap
mechanism already used for large variable-length attribute data):
\texttt{H5HG\_insert} for writing, \texttt{H5HG\_read} for reading, and
\texttt{free()} for closing. Plugins implement these callbacks only when
they need custom on-disk layout --- for example, alignment requirements
for GPU DMA transfers directly out of the file buffer.

The three callbacks are independent of \texttt{set\_config}/\texttt{get\_config}:
a filter's parameter string continues to carry small, human-readable
configuration, while the blob carries the large, opaque payload. A
filter that needs both derives any small values it needs from dataset
properties via the existing \texttt{set\_local} callback, or encodes
them as a short header inside the blob itself; the blob content is
entirely filter-defined bytes, so this requires no new library
mechanism.

\begin{minted}{c}
/* Fixed-size on-disk locator for a blob's global-heap object.
 * Opaque to plugin authors; passed back unchanged between write_blob
 * and read_blob. */
typedef struct H5Z_blob_loc_t {
    haddr_t addr;   /* global heap collection address */
    size_t  idx;    /* object index within the collection */
} H5Z_blob_loc_t;

/* Called at H5Dcreate time to persist a blob alongside the pipeline
 * message.  If NULL, the library uses its default H5HG-based writer.
 * loc_out identifies the stored object for the later read_blob call. */
typedef herr_t (*H5Z_write_blob_func_t)(hid_t            file_id,
                                        const void       *buf,
                                        size_t            size,
                                        H5Z_blob_loc_t   *loc_out);

/* Called at H5Dopen time to recover a blob written by write_blob.
 * If NULL, the library uses its default H5HG-based reader.
 * *buf_out is allocated by the callback (or the library, in the
 * default case); the caller releases it via close_blob. */
typedef herr_t (*H5Z_read_blob_func_t)(hid_t            file_id,
                                       H5Z_blob_loc_t    loc,
                                       void            **buf_out,
                                       size_t           *size_out);

/* Called at H5Dclose time to release the in-memory blob buffer
 * returned by read_blob.  If NULL, the library calls free(). */
typedef herr_t (*H5Z_close_blob_func_t)(void *buf, size_t size);
\end{minted}

Added to \texttt{H5Z\_class3\_t}:

\begin{minted}{c}
    H5Z_write_blob_func_t       write_blob;  /* NULL-able */
    H5Z_read_blob_func_t        read_blob;   /* NULL-able */
    H5Z_close_blob_func_t       close_blob;  /* NULL-able */
\end{minted}

\texttt{H5Z\_blob\_loc\_t} is an opaque locator handed unchanged from
\texttt{write\_blob} to the pipeline-message encoder and later from the
decoder to \texttt{read\_blob}; plugin authors do not construct or
interpret its fields.

Add \texttt{has\_blob\_callbacks} (a \texttt{bool}) to
\texttt{H5Z\_class\_info\_t} (RFC-HDFG-2026-001) so
\texttt{H5Zget\_filter\_class\_info} can report whether a registered
filter implements the blob callbacks.

\subsection{New Internal State: \texttt{H5Z\_filter\_info\_t} Additions}
\label{sec:blob-internal-state}

\begin{minted}{c}
/* H5Zprivate.h -- internal only; not part of the public API */
struct H5Z_filter_info_t {
    /* ... existing fields ... */
    void            *aux_data;  /* in-memory blob; NULL if this filter has none */
    size_t           aux_size;  /* byte length of aux_data                     */
    H5Z_blob_loc_t   aux_loc;   /* on-disk locator; undefined until written    */
};
\end{minted}


\section{On-Disk Format}
\label{sec:blob-format}

The baseline per-filter layout is the \texttt{H5O\_PLINE\_VERSION\_3}
record defined by RFC-HDFG-2026-001, which appends a verbatim
configuration-string block (\texttt{[config\_length:2][config:variable]})
to the \texttt{H5O\_PLINE\_VERSION\_2} fields:

\begin{center}
\texttt{[id:2][flags:2][cd\_nelmts:2][cd\_values:cd\_nelmts$\times$4][name:variable]}\\
\texttt{[config\_length:2][config:variable]}
\end{center}

This RFC's format adds a \texttt{has\_aux} bit after the
configuration-string block and, when set, the fixed-size
\texttt{H5Z\_blob\_loc\_t} encoding (address plus heap-object index, the
same fields \texttt{H5HG\_t} already carries internally for other
large-object references):

\begin{center}
\texttt{[id:2][flags:2][cd\_nelmts:2][cd\_values:cd\_nelmts$\times$4][name:variable]}\\
\texttt{[config\_length:2][config:variable]}\\
\texttt{[has\_aux:1][aux\_loc:variable if has\_aux]}
\end{center}

\textbf{Version number.} This RFC needs a pipeline-message format
bump. RFC-HDFG-2026-001 now introduces an on-disk format change of its
own and claims \texttt{H5O\_PLINE\_VERSION\_3} for the
configuration-string block.  Consequently:

\begin{itemize}
\item \textbf{If this RFC lands in the same release} as
  RFC-HDFG-2026-001's format work, the \texttt{has\_aux}/\texttt{aux\_loc}
  block is folded directly into \texttt{H5O\_PLINE\_VERSION\_3}, so the
  pipeline message takes a single bump (see ``one bump, not two,''
  \SectionRef{sec:release-targeting}).
\item \textbf{Otherwise}, this RFC claims
  \texttt{H5O\_PLINE\_VERSION\_4}, layered on the version-3 record as
  shown above --- unless some other RFC has claimed that number first by
  the time this RFC is finalized. Whoever finalizes this RFC should
  check for version-number conflicts with any other pipeline-message
  format work in flight at that time and claim the next free version.
\end{itemize}

The blob bytes themselves are stored as a global-heap (\texttt{H5HG})
object; the pipeline message records only the locator, not the bytes
inline. \texttt{H5O\_pline\_ver\_bounds[]} maps \texttt{H5F\_libver\_t}
to the pipeline message version; a DCPL with a blob-bearing filter
requires a libver upper bound that admits this RFC's pipeline-message
version, exactly as any other version-gated format feature does.

Runtime I/O is unaffected: \texttt{H5Z\_pipeline}, the hot path for
chunk I/O, touches none of this. The blob is configuration data, loaded
once at open time; compression/decompression performance is unchanged.

\subsection{Alternative Considered: No-Format-Change Attribute Convention}
\label{sec:blob-attr-fallback}

A file format change is unavoidable for \emph{first-class} blob
support: every field in the current pipeline message record is a small
fixed integer or the inline name/\texttt{cd\_values}; the global-heap
locator has nowhere to live. The one genuine no-format-change
alternative is a reserved-attribute convention --- store the blob (or
its heap locator) in a conventionally-named attribute on the dataset,
the way dimension scales are implemented entirely as
\texttt{DIMENSION\_LIST}/\texttt{REFERENCE\_LIST} attributes with no
format change.

This RFC \textbf{rejects} that fallback for the primary design, for
concrete reasons:

\begin{itemize}
\item \textbf{No structural binding to a pipeline entry.} An attribute
  hangs off the dataset, not off filter entry $i$ of the pipeline; the
  association (including the duplicate-filter-ID case,
  \SectionRef{sec:append-filter-blob}) would be by fragile naming
  convention.
\item \textbf{No lifecycle coupling.} The pipeline message's
  copy/delete paths (\texttt{H5O\_\_pline\_copy\_file}, the
  \texttt{delete} handler) would not carry or reclaim the attribute;
  every tool and code path that touches pipelines would need parallel
  attribute plumbing.
\item \textbf{Namespace pollution.} The blob becomes a user-visible,
  user-mutable attribute, with collision risk against application
  attribute names.
\end{itemize}

It remains available as a stopgap a filter author can implement
privately today (a filter's \texttt{set\_local} can read a
conventionally-named attribute via the dataset's properties), and
nothing in this RFC forbids that --- but the library-supported
mechanism is the versioned pipeline message.

\subsection{Release and Format-Generation Targeting}
\label{sec:release-targeting}

The pipeline-message version mechanism does not require a major
release. The standard, repeatedly-used pattern is: add the new message
version, add a row to \texttt{H5O\_pline\_ver\_bounds[]}, and let
\texttt{H5Pset\_libver\_bounds} gate emission --- exactly how the
layout message v4 (virtual datasets) shipped in HDF5~1.10, and how the
datatype, fill-value, and link-info messages have all evolved. Older
libraries encountering the new version fail cleanly with ``bad version
number for filter pipeline message.''

Two coordination requirements, rather than technical blockers, govern
the targeting decision:

\begin{itemize}
\item \textbf{Same-release constraint} (\SectionRef{sec:relationship}):
  the filter-class callbacks and the pipeline-message version must ship
  together.
\item \textbf{One bump, not two.} RFC-HDFG-2026-001 now defines
  \texttt{H5O\_PLINE\_VERSION\_3}, which persists the filter's
  configuration string verbatim in the pipeline message. (Its earlier
  draft instead deferred a typed \texttt{cd\_values} type-tag extension
  to future work; that design is superseded and will not consume a
  version number.) If this RFC lands within the same planning horizon,
  the blob locator should be folded into that \emph{same} version~3
  record and the same format bound, so the pipeline message does not
  take two consecutive incompatible bumps; otherwise this RFC bumps to
  version~4 (\SectionRef{sec:blob-format}).
\end{itemize}

If the project designates a format-generation release (a ``3.0'' in the
reviewer's framing) for accumulated format work, this RFC's message
version should be introduced at that release's format bound; note that
\texttt{H5F\_LIBVER\_V200} already exists as a defined format slot at
which no message currently bumps its version, illustrating that bounds
and message versions are decoupled. This RFC takes no position on
release numbering; it requires only libver gating and the two
coordination rules above.


\section{Public API}
\label{sec:public-api}

\subsection{\texttt{H5Pappend\_filter\_blob}}
\label{sec:append-filter-blob}

\begin{minted}{c}
herr_t H5Pappend_filter_blob(
    hid_t         plist_id,
    H5Z_filter_t  id,
    unsigned      flags,
    const void   *buf,
    size_t        size
);
\end{minted}

\verysubsection{Description}
Appends filter \texttt{id} to the pipeline on \texttt{plist\_id}, exactly
as RFC-HDFG-2026-001's \texttt{H5Pappend\_filter} does, and attaches
\texttt{buf}/\texttt{size} as that entry's blob. The bytes are opaque to
the library; they are handed to the filter's \texttt{write\_blob}
callback (or the default \texttt{H5HG} writer) at \texttt{H5Dcreate}
time and recovered via \texttt{read\_blob} (or the default reader) at
\texttt{H5Dopen} time. \texttt{set\_config} is not invoked for this entry
point --- the blob path bypasses the parameter-string layer entirely,
since the whole point of the blob channel is to carry data that does
not fit through it.

\textbf{The blob is associated with the pipeline \emph{entry}, not with
the filter ID.} HDF5 pipelines legally contain multiple entries with
the same filter ID, and the existing ID-addressed APIs
(\texttt{H5Pmodify\_filter}, \texttt{H5Z\_filter\_in\_pline}) resolve an
ID to the \emph{first} matching entry. A blob keyed by ID would inherit
that first-match ambiguity. \texttt{H5Pappend\_filter\_blob} avoids it
structurally: each call creates one new pipeline entry and binds the
blob to that entry (positionally, in \texttt{H5Z\_filter\_info\_t});
two entries with the same ID and different blobs are well-defined. Any
future blob accessor (\SectionRef{sec:blob-getter}) must likewise
address entries by pipeline index --- as \texttt{H5Pget\_filter2} does
--- never by filter ID.

\textbf{Decision: single entry point, no \texttt{H5Z\_params\_t}
variant.} An earlier draft considered adding \texttt{H5Z\_PARAMS\_BLOB}
as a third \texttt{H5Z\_params\_type\_t} tag so that blob configuration
could flow through \texttt{H5Pappend\_filter}'s tagged union. This RFC
does not do so: \texttt{H5Z\_params\_t} is designed around values that
fit comfortably in a compound literal, and forcing a multi-megabyte
buffer through that union would either double-store the buffer or
complicate its lifetime rules for no benefit. \texttt{H5Pappend\_filter\_blob}
is a dedicated function with a signature suited to its purpose;
\texttt{H5Z\_params\_type\_t} is unchanged by this RFC.

\verysubsection{Parameters}
\begin{itemize}
\item \texttt{plist\_id}---A dataset creation property list identifier.
\item \texttt{id}---The filter identifier. The registered filter must
  implement \texttt{write\_blob}/\texttt{read\_blob} or accept the
  library's default \texttt{H5HG}-based implementation.
\item \texttt{flags}---Same semantics as \texttt{H5Pappend\_filter}.
\item \texttt{buf}---Pointer to the blob bytes. The library copies these
  bytes into DCPL-owned storage before returning (\SectionRef{sec:blob-encode-decode});
  the caller's buffer may be freed or reused immediately after the call.
\item \texttt{size}---Length of \texttt{buf} in bytes. There is no fixed
  upper bound analogous to \texttt{H5Z\_CONFIG\_STRING\_MAX}; the
  practical limit is available memory (note that the blob is deep-copied
  with the DCPL, \SectionRef{sec:blob-per-dataset}).
\end{itemize}

\verysubsection{Returns}
Non-negative on success; negative on failure.

\verysubsection{Errors}
\begin{itemize}
\item \texttt{H5E\_NOFILTER}---The filter plugin could not be found or
  loaded.
\item \texttt{H5E\_BADVALUE}---\texttt{buf} is \texttt{NULL} with nonzero
  \texttt{size}, or \texttt{size} is zero with non-NULL \texttt{buf}.
\end{itemize}

\subsection{Programming Model: Per-Dataset Blobs}
\label{sec:blob-per-dataset}

A DCPL is a template: at \texttt{H5Dcreate} the dataset takes a private
deep copy of it (\texttt{H5P\_copy\_plist} in \texttt{H5D\_\_create}),
so mutating the caller's DCPL never affects an existing dataset. A
direct consequence is the question every reviewer asks first:
\emph{``one DCPL, two datasets, same filter, different blobs?''}

\noindent\textbf{Answer: one DCPL per distinct blob} (create a fresh
DCPL, or \texttt{H5Pcopy} an existing one and re-append the filter with
the other blob). This is not a workaround; it is the established HDF5
idiom for per-dataset configuration that lives in the DCPL:

\begin{itemize}
\item \textbf{Virtual dataset mappings} (\texttt{H5Pset\_virtual}) ---
  inherently per-dataset source-file/dataset/selection mappings, stored
  in the DCPL layout property; one DCPL per VDS.
\item \textbf{External file lists} (\texttt{H5Pset\_external}) ---
  per-dataset file names in the DCPL; distinct configuration, distinct
  DCPL.
\item \texttt{h5repack} itself uses copy-then-tweak
  (\texttt{H5Pcopy(dcpl\_in)} then adjust the pipeline) for every output
  dataset.
\end{itemize}

\verysubsection{Cost note: the blob is deep-copied with the pipeline}

The pipeline property is deep-copied on every \texttt{H5Pcopy}
\emph{and} on every \texttt{H5Dcreate} (each dataset's private DCPL
copy). A multi-megabyte blob therefore gets duplicated on each of these
operations under the na\"ive implementation. The implementation should
store \texttt{aux\_data} in a reference-counted (copy-on-write) buffer
inside the pipeline property so DCPL copies share one allocation; the
buffer is immutable after \texttt{H5Pappend\_filter\_blob}, so sharing
is safe. This is an implementation-quality requirement, not a format or
API question.

\verysubsection{Alternatives considered for per-dataset association}

\begin{itemize}
\item \textbf{Create-time argument} (blob passed to a new
  \texttt{H5Dcreate}-family call alongside \texttt{type\_id}/\texttt{space\_id},
  which are precedent for per-dataset arguments): keeps the DCPL a pure
  reusable template, but adds a create-signature variant --- the very
  thing property lists exist to avoid. Rejected for now; revisit if the
  one-DCPL-per-blob model proves burdensome in practice.
\item \textbf{Post-create attach} (an \texttt{H5Dset\_extent}-style
  operation on an existing dataset): keeps the template clean, but the
  blob would not exist at \texttt{H5Dcreate} time --- after the pipeline
  message is first written --- forcing an object-header rewrite and
  ordering/atomicity questions. Rejected.
\item \textbf{\texttt{set\_local}-driven override}: \texttt{H5Z\_set\_local}
  already runs against the dataset's \emph{private} DCPL copy, and
  built-in filters (shuffle, szip) already rewrite their own pipeline
  entry per dataset there. A filter whose blob is \emph{derivable} from
  the dataset's datatype/dataspace can compute and install it in
  \texttt{set\_local}, giving true one-template-DCPL behavior with no
  new API. This works today under this RFC's design and is the
  recommended pattern for author-computed blobs; it does not help
  \emph{caller-supplied} per-dataset blobs, which use the
  one-DCPL-per-blob model above.
\end{itemize}


\section{Dataset Create / Open / Close Lifecycle}
\label{sec:blob-lifecycle}

\textbf{Create} (\texttt{H5D\_\_create}), after \texttt{H5Z\_set\_local}
runs, for each filter with \texttt{aux\_data~!=~NULL}:
\begin{enumerate}
\item Call \texttt{filter->cls->write\_blob(file\_id, aux\_data, aux\_size,
  \&aux\_loc)}, or the default \texttt{H5HG} writer if
  \texttt{write\_blob} is \texttt{NULL}.
\item Store \texttt{aux\_loc} in \texttt{H5Z\_filter\_info\_t}.
\item Encode the pipeline message at this RFC's format version with
  \texttt{has\_aux} set and \texttt{aux\_loc} populated.
\end{enumerate}

\textbf{Open} (\texttt{H5D\_\_open}):
\begin{enumerate}
\item Decode the pipeline message; for each filter with \texttt{has\_aux},
  read \texttt{aux\_loc}.
\item Call \texttt{filter->cls->read\_blob(file\_id, aux\_loc, \&aux\_data,
  \&aux\_size)}, or the default \texttt{H5HG} reader if
  \texttt{read\_blob} is \texttt{NULL}.
\item Populate \texttt{aux\_data}/\texttt{aux\_size} in
  \texttt{H5Z\_filter\_info\_t}.
\end{enumerate}

\textbf{Close}: Call \texttt{close\_blob(aux\_data, aux\_size)}, or
\texttt{free()} if \texttt{close\_blob} is \texttt{NULL}, and clear the
fields.

\textbf{Delete} (dataset deleted, or the filter removed from a pipeline
being modified): the pipeline message's \texttt{delete} handler in
\texttt{H5Opline.c} calls \texttt{H5HG\_remove} on \texttt{aux\_loc} to
reclaim the heap space, analogous to how the fill-value message's
\texttt{delete} handler frees vlen data today (\texttt{H5Ofill.c}).


\section{Design Decisions}
\label{sec:decisions}

Of the design questions originally raised for this feature, three ---
the \texttt{H5Pencode}/\texttt{H5Pdecode} policy, \texttt{h5repack}
behavior, and the storage format --- are resolved below. The parallel
I/O write protocol (\SectionRef{sec:blob-parallel}) is \emph{proposed
but pending prototype validation}: review against the library's
parallel metadata-cache design invalidated the protocol an earlier
draft had recorded as resolved. The reference-to-dataset use case
(\SectionRef{sec:blob-usecaseb}) was likewise redesigned following
review.

\subsection{Parallel I/O Write Protocol (Proposed, Pending Validation)}
\label{sec:blob-parallel}

\begin{underdiscussion}
This subsection records a proposed protocol, not a resolved decision.
An earlier draft resolved this question as ``rank~0 writes the blob and
broadcasts its locator''; checking that protocol against the parallel
metadata-cache implementation showed it to be unworkable (see below).
The replacement must be validated by a prototype, reviewed with the
metadata-cache maintainers, before this subsection can be marked
resolved.
\end{underdiscussion}

\verysubsection{Why rank-0-only writing fails}

\texttt{H5HG\_insert} performs three metadata-cache-visible actions:
a file-space allocation (\texttt{H5MF\_alloc}, \texttt{H5HG.c}), a
cache-entry insert (\texttt{H5AC\_insert\_entry}), and a
dirty-on-unprotect (\texttt{H5AC\_\_DIRTIED\_FLAG}). The parallel
metadata cache's coherency protocol requires that \emph{every rank
perform the identical sequence of metadata operations}; the invariant
is stated in \texttt{H5ACmpio.c}: \emph{``all operations writing
metadata must be collective. Thus all metadata caches see the same
sequence of operations, and therefore the same dirty data creation.''}
The mechanism that enforces it is concrete: each rank advances a
\texttt{dirty\_bytes} counter for every insert/dirty, and collective
cache sync points trigger when that counter crosses a shared threshold.
If only rank~0 inserted the blob's heap object, rank~0's counter and
cache contents would diverge from every other rank's, desynchronizing
the collective sync points --- deadlock or corruption. This is
precisely why variable-length and region-reference \emph{writes} are
disabled outright under MPI (\texttt{H5Dio.c}: \emph{``they use the
global heap in the file and we don't support parallel access of that
yet''}) rather than special-cased. The rank-0-plus-broadcast patterns
that do exist in the library (superblock-signature search, collective
read optimizations) are read-side and raw/VFD-level; no code today
broadcasts a newly created dirty metadata entry, and no machinery
exists to inject one into other ranks' caches.

\verysubsection{Proposed protocol: collective identical insert, no broadcast}

At \texttt{H5Dcreate} time the situation is better than the general
global-heap-write problem, because dataset-creation metadata in
parallel HDF5 is already produced \emph{redundantly but identically on
every rank}: each rank runs the same code against the same cache state,
and the cache layer coordinates who physically writes. A blob supplied
via \texttt{H5Pappend\_filter\_blob} is part of the DCPL, and parallel
HDF5 already requires the DCPL to be identical on all ranks at a
collective \texttt{H5Dcreate}. The proposal is therefore:

\begin{itemize}
\item \textbf{Every rank calls the identical default blob writer}
  (\texttt{H5HG\_insert} with identical bytes and size) at the same
  point in \texttt{H5D\_\_create}. Under the identical-operations
  invariant, every rank computes the same \texttt{H5MF\_alloc} address
  from identical free-space state and inserts the identical dirty
  entry; every rank therefore arrives at the same
  \texttt{H5Z\_blob\_loc\_t} deterministically, and \textbf{no broadcast
  is needed at all}.
\item This does not conflict with the VL/region-reference write ban:
  that ban concerns \texttt{H5Dwrite}-time element data, which differs
  per rank and flows through independent I/O paths. The blob is
  create-time, collective, and rank-identical by DCPL rules --- exactly
  the conditions the redundant-metadata model requires.
\end{itemize}

\verysubsection{Open items before this can be marked resolved}

\begin{itemize}
\item The determinism argument inherits the identical-operations
  invariant; any \emph{pre-existing} rank divergence would surface here
  as silent locator divergence. A debug-build consistency check (e.g.,
  \texttt{MPI\_Bcast} rank~0's locator and compare) should be part of
  the prototype.
\item Interaction with paged allocation and free-space-manager
  settings, and the \texttt{H5FD\_MEM\_GHEAP} allocation type, needs
  checking.
\item Validation and review with the parallel metadata-cache
  maintainers.
\end{itemize}

\verysubsection{Custom callbacks in parallel (normative)}

A filter's custom \texttt{write\_blob} inherits the same requirement:
in a parallel job it \textbf{must} perform identical file-modifying
operations on every rank (or perform none, delegating storage to the
default writer). A \texttt{write\_blob} that behaves differently per
rank --- including one that internally attempts rank-0-only writing ---
is undefined behavior in parallel. The library documentation for
\texttt{H5Z\_write\_blob\_func\_t} must state this.

\subsection{\texttt{H5Pencode}/\texttt{H5Pdecode} Policy}
\label{sec:blob-encode-decode}

\noindent\textbf{Decision:} \texttt{H5Pencode} serializes the blob's
bytes inline; \texttt{H5Pdecode} leaves the on-disk locator undefined
until the next \texttt{H5Dcreate}.

\texttt{H5P\_\_ocrt\_pipeline\_enc} encodes filter ID, flags, name, and
\texttt{cd\_values} --- all in-memory data; no file address appears in
an encoded DCPL. Fill-value encoding follows the same rule: a serialized
property must be self-contained, able to survive a buffer, a process
boundary, or a decode without the original file. \texttt{aux\_loc} is a
derived, file-specific value; \texttt{aux\_data} is the canonical
in-memory representation. \texttt{H5P\_\_ocrt\_pipeline\_enc} is
extended to append, after each filter's \texttt{cd\_values}:

\begin{center}
\texttt{[has\_aux:1][aux\_size:8 if has\_aux][aux\_data:aux\_size bytes if has\_aux]}
\end{center}

\texttt{H5P\_\_ocrt\_pipeline\_dec} reads those fields back into
\texttt{aux\_data}/\texttt{aux\_size} and leaves \texttt{aux\_loc}
undefined; a subsequent \texttt{H5D\_\_create} using the decoded DCPL
triggers \texttt{write\_blob} and assigns a real locator in whichever
file it is used with.

Serializing the locator instead of the bytes was rejected: it would make
an encoded DCPL valid only in its originating file, breaking every
\texttt{H5Pdecode}-then-\texttt{H5Dcreate} workflow in a different file.
Returning an error for blob-bearing DCPLs was also rejected as
unnecessarily restrictive, since \texttt{aux\_data} is always available
in memory after \texttt{H5D\_\_open} or after
\texttt{H5Pappend\_filter\_blob}. The tradeoff accepted is that encoded
DCPLs carrying a multi-megabyte blob are correspondingly large; this is
acceptable since \texttt{H5Pencode} is not a hot path and callers
invoking it on such a DCPL already know they hold unusual data.

\subsection{\texttt{h5repack} and \texttt{H5Pcopy} Behavior}
\label{sec:blob-repack}

\noindent\textbf{Decision:} No \texttt{h5repack}-specific code is
needed; \texttt{H5Pcopy} is extended to deep-copy \texttt{aux\_data}.

\texttt{h5repack} copies a normal (non-external) dataset's DCPL via
\texttt{H5Pcopy(dcpl\_in)} and does not call \texttt{H5HG} functions
directly. The existing lifecycle already makes this work automatically,
\emph{provided} \texttt{H5Pcopy} performs a deep copy of
\texttt{aux\_data}:

\begin{enumerate}
\item \texttt{H5Dopen} on the source populates \texttt{aux\_data} in the
  source DCPL's filter entry (via \texttt{read\_blob} or the default
  reader).
\item \texttt{H5Pcopy(dcpl\_in)} must give the copy its own logical
  \texttt{aux\_data} (a deep copy, or a shared reference-counted buffer
  per \SectionRef{sec:blob-per-dataset}), leaving \texttt{aux\_loc}
  undefined in the copy (the destination address is not yet known).
\item \texttt{H5Dcreate} with the copied DCPL in the destination file
  triggers \texttt{write\_blob} (or the default writer), producing a new
  heap object at a new locator in the destination file. Only the bytes
  travel; the source locator is never copied.
\end{enumerate}

The pipeline message's \texttt{copy} callback in \texttt{H5Opline.c} is
extended accordingly. When a filter is removed from a pipeline (e.g.,
\texttt{h5repack}'s \texttt{apply\_filters()} replacing a filter), the
pipeline message's destructor frees \texttt{aux\_data} the same way it
already must free other per-filter allocations.

\subsection{Storage Format: Naked \texttt{H5HG}}
\label{sec:blob-storage-format}

\noindent\textbf{Decision:} Blob bytes are stored as a plain
\texttt{H5HG} global-heap object, not a new HDF5 object message type.

A new message type (\texttt{H5O\_MSG\_FILTER\_BLOB}) would let
\texttt{h5dump} label the object (e.g., ``4~MB of SZ4 JIT source''), but
the payload remains filter-plugin-specific binary data that no generic
tool can render meaningfully --- the pipeline message already identifies
the filter, which is the useful part of that label. A new message type
also carries an ongoing maintenance cost (roughly a dozen callbacks:
encode, decode, copy, \texttt{raw\_size}, reset, free, delete, link, and
the pre/copy/post-copy-file hooks) for marginal tooling value. Existing
HDF5 precedent for ``large binary thing referenced from the object
header, recovered on demand'' is exactly \texttt{H5HG}, already used for
large variable-length attribute data. If a genuine tooling need for a
self-describing message type emerges after deployment, one can be
introduced in a later file-format version without breaking files
written under this design --- the pipeline message's \texttt{aux\_loc}
field would simply become one of several supported locator kinds.

\textbf{\texttt{del} callback obligation:} When a pipeline message is
deleted (dataset deleted, filter removed), \texttt{H5HG\_remove(file, \&loc)}
must be called to reclaim the heap space. This is implemented in the
pipeline message \texttt{delete} handler in \texttt{H5Opline.c},
analogous to how the fill-value message's \texttt{delete} handler frees
vlen data (\texttt{H5Ofill.c}).

\subsection{Use Case B: Reference to Another Dataset (Revised)}
\label{sec:blob-usecaseb}

Use Case B (ROIBIN-SZ) needs the filter to locate a user-visible mask
dataset at open time. An earlier draft resolved this by having the
filter store an \texttt{H5R\_ref\_t} object reference as its blob and
dereference it inside \texttt{read\_blob} using the supplied
\texttt{file\_id}. \textbf{That design is withdrawn}: it cannot be
implemented correctly by a plugin today.

\verysubsection{Why raw reference bytes do not persist}

The public \texttt{H5R\_ref\_t} is a 64-byte opaque buffer whose actual
contents (\texttt{H5R\_ref\_priv\_t}) include a \emph{cached process-local}
\texttt{hid\_t loc\_id}, an application-reference-count flag, and ---
depending on reference type --- raw heap pointers (\texttt{char
*filename}, \texttt{H5S\_t *space}, \texttt{char *name}). Copied to
disk and read back in a later session, the stale \texttt{loc\_id} and
dangling pointers make any use of the bytes undefined. The library
itself never persists this form: when a reference is written into a
dataset or attribute, the reference-datatype conversion path calls the
\emph{internal} encoder \texttt{H5R\_\_encode} (versioned header,
length-prefixed object token, serialized filename/selection/name) and
stores the encoded stream through the global-heap blob layer; on read,
\texttt{H5R\_\_decode} deliberately resets \texttt{loc\_id} to
\texttt{H5I\_INVALID\_HID} and the library re-attaches the currently
open file. Crucially, \texttt{H5R\_\_encode}/\texttt{H5R\_\_decode} are
package-private --- \textbf{there is no public
\texttt{H5Rencode}/\texttt{H5Rdecode}} --- so a filter plugin has no
supported way to produce or consume the persistable encoding itself.

\verysubsection{Revised design (primary): store the mask dataset's path}

The filter stores the mask dataset's absolute path name inside its blob
(the blob is filter-defined bytes; a NUL-terminated path plus any
filter-private header is an ordinary payload). At \texttt{H5Dopen} time,
\texttt{read\_blob} --- or the filter's first use --- opens the mask via
\texttt{H5Dopen2(file\_id, path, H5P\_DEFAULT)} using the
\texttt{file\_id} the callback already receives. This requires no
reference machinery, no extra global-heap object, no temporary
reference objects, and no new public API; and because the path names an
object in the same file, \texttt{h5repack} copies (which copy the whole
file's object tree) keep it valid. \textbf{Limitation (documented, not
solved):} a path is not rename-proof --- moving or renaming the mask
dataset breaks the association. Applications that rename must update
the blob (recreate the dataset or re-append the filter).

\verysubsection{Alternative (prerequisite API): public
\texttt{H5Rencode}/\texttt{H5Rdecode}}

If rename-proof semantics are required, the clean path is a small,
separate RFC adding public \texttt{H5Rencode}/\texttt{H5Rdecode}
wrappers over the existing internal encoder (the same
size-query-then-fill convention as \texttt{H5Sencode}/\texttt{H5Tencode}/
\texttt{H5Pencode}). A filter would then store the \emph{encoded}
reference in its blob, call \texttt{H5Rdecode} in \texttt{read\_blob},
and open the object via \texttt{H5Ropen\_object} with the supplied
\texttt{file\_id} as the location. This RFC does not gate on that
addition; it names it as the supported future upgrade path from
path-string association.

\verysubsection{Partial I/O on the mask}

Reading the entire mask into memory inside \texttt{read\_blob} at
\texttt{H5Dopen} time is the simple model but forecloses partial I/O
for very large masks. Two conforming alternatives: the filter may hold
the opened \texttt{hid\_t} for the mask dataset and defer actual reads
to filter-execution time, using the chunk coordinates
(\texttt{scaled}/\texttt{ndims}) that \texttt{H5Z\_func2\_t}
(RFC-HDFG-2026-001) supplies to select and read only the mask region
covering the current chunk; or, for version~1 of a filter, accept
full-mask materialization at open and document the memory cost. The
library imposes no policy here; the blob carries only the path.


\section{Under Discussion: Retrieving a Default-Stored Blob}
\label{sec:blob-getter}

\begin{underdiscussion}
No decision is recorded for this section; it exists to solicit review
feedback before an accessor is added to the implementation.
\end{underdiscussion}

Prototype experience surfaced a gap in the default-storage path: a blob
written through the default \texttt{H5HG} writer is currently
\emph{write-only}. The library recovers the bytes into the pipeline at
\texttt{H5Dopen}, and they propagate correctly through \texttt{H5Pcopy},
\texttt{H5Pencode}/\texttt{H5Pdecode}, and \texttt{h5repack} --- but no
public API exposes them. An application cannot read a blob back from a
dataset's DCPL, and a filter cannot consume its own blob unless it
implements \emph{both} \texttt{write\_blob} and \texttt{read\_blob}: a
class that implements only \texttt{read\_blob} cannot interpret the
default writer's locator, because \texttt{H5HG} is library-private. Both
use cases in \SectionRef{sec:motivation} therefore require custom
callbacks today, which makes the default storage path considerably less
useful than it appears.

\noindent\textbf{Option under discussion:} a public accessor,

\begin{minted}{c}
herr_t H5Pget_filter_blob(
    hid_t     plist_id,
    unsigned  idx,        /* pipeline index, as in H5Pget_filter2       */
    void     *buf,        /* NULL for a size query                      */
    size_t   *size        /* in: capacity of buf; out: blob byte count */
);
\end{minted}

\noindent following the same size-query convention as
RFC-HDFG-2026-001's \texttt{H5Pget\_filter\_params\_by\_idx}. This would
let applications and tools (e.g., an \texttt{h5dump~-p} annotation that
a blob is present and its size) retrieve default-stored blobs, and ---
notably --- would let a \emph{filter} consume its blob without
implementing any storage callbacks: the \texttt{set\_local} callback
already receives the dataset's \texttt{dcpl\_id} and could fetch and
cache the blob there. That combination (default storage +
\texttt{set\_local} retrieval) would cover Use Case A with no
filter-side storage code at all.

Alternatives considered but not currently favoured: keeping the status
quo (filters that consume blobs must implement both storage callbacks),
or passing the blob into the filter callback through a new context
parameter (rejected for widening the hot-path signature).

\end{document}
